---
title: "Assignment 2: Part II (version 2)"
subtitle: "MAT-32806 Statistics for Data Scientists"
author: 
  - Citlali Melchor Ramírez^[Reg. No. 930522157110, Biosystems Engineering, <citlali.melchorramirez@wur.nl>]
date: "`r format(Sys.time(), '%d %b %Y')`"
output:
  bookdown::pdf_document2
geometry: left=1in, right=1in, top=1in, bottom=1in
classoption: a4paper
spacing: single
bibliography: genome.bib
link-citations: true

---


```{r setup, include=FALSE}
# Chunks configuration 
knitr::opts_chunk$set(echo = FALSE, 
                      results='hide',
                      warning=FALSE, 
                      message=FALSE,
                      fig.width=4, 
                      fig.height=3, 
                      tidy = FALSE)
# Libraries used in this assignment
library (pander)
library (gam)
library (mgcv)
library (leaps)
library (ISLR)
```

\pagebreak 

# Introduction
\vspace{-3truemm}
It is wanted to predict grain number in a specific maize hybrid across 25 different environments, using a number of environmental characterizations: _Average minimum temperature during the night one of three growth periods_ ($T_{night.1},\ T_{night.2},\ T_{night.3}$), _Average radiation intercepted during the same three periods_ ($R_{i.1},\ R_{i.2},\ R_{i.3}$), _Average soil water potential_ ($P_{si.2},\ P_{si.3}$) and _Average maximum temperature_ ($T_{max.1},\ T_{max.2},\ T_{max.3}$). @Millet749 fitted a simple linear model for the full set of hybrids with the predictors $T_{min.2}$, $R_{i.1}$, and $P_{si.2}$. The aim of this practical is to try out different models to see whether it is possible to improve this simple model, which only uses 3 degrees of freedom. 
\vspace{-3truemm}

```{r Data2, tidy=TRUE}
# Import the data
Data2<-read.csv('maize data SDS assignment v2.csv',header=T)
attach(Data2)
```

# Generalized Additive Models 
\vspace{-3truemm}
The _Generalized Additive Models_ (GAM) are extensions to standard linear models allowing non-linear functions and maintaining additivity. 
\vspace{-4truemm}

## Original model 
\vspace{-3truemm}
The first model (`Linear`) is the model from @Millet749, linear in all three predictors. Table \@ref(tab:model1) shows the ANOVA analysis of this model; we can see that the p-values for the predictors are smaller than 0.05 which means they are statistically significant. In Figure \@ref(fig:originalModelplot) we can see that the number of grains decreases with the minimum night temperature and increases with the radiation intercepted and the soil water potential. 
\vspace{-4truemm}

## Polynomial 
\vspace{-3truemm}
The next model analysed was the case in which a second degree polynomial is added (`Polynomial`). Table \@ref(tab:model2) shows the ANOVA analysis of this model, where we can see that the predictors are significant; moreover, in Figure \@ref(fig:model2Plot) we can see that the relation remains similar with the previous model for $T_{night.w}$ and $P_{si.2}$, but changes for the radiation, where the grains number seems to reach a maximum around 280 and then starts to decrease. 
\vspace{-4truemm}

## Natural Splines
\vspace{-3truemm}
Then, it was considered a model with natural splines of 4 degrees of freedom (`Gamns`). Table \@ref(tab:model3) shows the ANOVA analysis of this model; in contrast with the previous models, the predictor $P_{si.2}$ presents a value slightly higher than 0.05. Moreover in Figure \@ref(fig:model3Plot) we can see that the relations between predictors and response is not as simple to interpret as with the previous models.
\vspace{-4truemm}

## Combined functions
\vspace{-3truemm}
Another possibility with GAM models is to combine different functions for each predictor (`GamMix`). An example of this was implemented in a mixed model, where the functions given to each predictor were linear, logarithmic and natural spline, for $T_{min.2}$, $R_{i.1}$, and $P_{si.2}$, respectively.  Table \@ref(tab:model4) shows the ANOVA analysis of this model and Figure \@ref(fig:model4Plot) the corresponding plots. 
\vspace{-4truemm}

## Different predictors: _Subset selection_
\vspace{-3truemm}
Finally, it was consider the possibility of having a different subset of predictors than those used in the original model. To analyse this, it was used the `regsubset` function with the `backward` method (Table \@ref(tab:SubSelec)), and the selected predictors were $T_{max.1}$, $T_{night.2}$, $P_{si.2}$. It was chosen a Natural Splines Model with 4 degrees of freedom (`GamnsDiff`). It is important to mention that the number of predictors stuck to three, to maintain the simplicity of the model, even thought a higher number of predictors could improve even more the model performance (Figure \@ref(fig:predSelect)). Table \@ref(tab:model5) shows the ANOVA analysis for this model. 
\vspace{-3truemm}

# Conclusions
\vspace{-3truemm}
Table \@ref(tab:summ) shows a summary with the $R^2-adjusted$ values, the _Generalized Cross Validation_ errors (GCV) and the _Scale estimate_ for the above mentioned models. It is possible to observe that the Natural Splines models (`Gamns` and `GamnsDiff`) have a lower squared residual standard error , which indicates an improvement over the original model. However,the `Gamns` model has a higher GCV value than the original Linear model. Furthermore, the `GamnsDiff` model, which uses different predictors, shows a lower GCV error and Scale estimate in comparison with the rest of the models. From this results it is possible to conclude that the `GamnsDiff` model performs the best, from the statistical point of view because it is more flexible, nevertheless there is a trade-off with its interpretability.

```{r Allmodels}
# Original model 
fit.lin <- gam(grain.number ~ 
                 Tnight.2 + 
                 Ri.1 + 
                 Psi.2)
# Second degree polynomial model
fit.poly <- gam(grain.number ~ 
                  poly(Tnight.2,degree=2) + 
                  poly(Ri.1, degree=2) + 
                  poly(Psi.2, degree=2))
# Natural splines model
fit.gamns <- gam(grain.number ~ 
                   ns(Tnight.2,df = 4) + 
                   ns(Ri.1, df = 4) + 
                   ns(Psi.2, df=4))
# Combined functions model
fit.gammix <- gam(grain.number ~ 
                  Tnight.2  + 
                  lo(Ri.1, span=0.7) + 
                  ns(Psi.2, df=4))
#Different predictors
fit.gamnsDiff <- gam(grain.number ~
                   ns(Tmax.1, df = 4) +
                   ns(Tnight.2, df=4) +
                   ns(Psi.2, df=4))
```



# References

<div id="refs"></div>

# Appendices
## Tables


```{r tab:model1, results='asis'}
# Original model
pander(anova(fit.lin)$pTerms.table, 
       caption='(#tab:model1)ANOVA for the original model from @Millet749')
```


```{r tab:model2, results='asis'}
# Second degree polynomial model
pander(anova(fit.poly)$pTerms.table, 
       caption='(#tab:model2)ANOVA for second degree polynomial model' )
```

```{r tab:model3, results='asis'}
# Natural splines model
pander(anova(fit.gamns)$pTerms.table,
       caption='(#tab:model3)ANOVA for natural splines model' )
```

```{r tab:model4, results='asis'}
# Combined functions model
pander(anova(fit.gammix)$pTerms.table,
       caption='(#tab:model4)ANOVA for combined functions model' )
```

```{r tab:model5, results='asis'}
# Different predictors
pander(anova(fit.gamnsDiff)$pTerms.table,
       caption="(#tab:model5) ANOVA for natural splines 
       model with different predictors ")
```

```{r tab:SubSelec, results='asis'}
# Subset selection 
regfit.full=regsubsets(grain.number~
                         Tnight.1 + Tnight.2 + Tnight.3 +
                         Ri.1 + Ri.2 + Ri.3 +
                         Psi.2 + Psi.3 +
                         Tmax.1 + Tmax.2 + Tmax.3,
                       nvmax=10,
                       data=Data2,
                       method="backward")
pander(summary(regfit.full)$outmat, 
       caption="(#tab:SubSelec) Subset selection of predictors", split.table=Inf)
```

```{r tab:summ, results='asis'} 
Rsq<-list ('names'=c('Linear','Polynomial', 'Gamns', 'GamMix','GamnsDiff'), 
               'R2-adjusted'=
             c(summary(fit.lin)$r.sq,
                  summary(fit.poly)$r.sq,
                  summary(fit.gamns)$r.sq,
                  summary(fit.gammix)$r.sq,
                  summary(fit.gamnsDiff)$r.sq),
           'Generalized Cross Validation'=
             format(c(summary(fit.lin)$sp.criterion, 
                  summary(fit.poly)$sp.criterion, 
                  summary(fit.gamns)$sp.criterion,
                  summary(fit.gammix)$sp.criterion,
                   summary(fit.gamnsDiff)$sp.criterion), scientific=TRUE, digits=4),
           'Squared Residual Standard Error'= 
             format(c(summary(fit.lin)$scale,
                  summary(fit.poly)$scale, 
                  summary(fit.gamns)$scale,
                  summary(fit.gammix)$scale,
                  summary(fit.gamnsDiff)$scale), scientific=TRUE, digits=4)
             )
pander(as.data.frame(Rsq, col.names = c('Model','R2-adjusted', 'GCV','Scale-est.' )), 
       caption='(#tab:summ) Models Summary')
```


## Figures

```{r originalModelplot, fig.width=8, fig.cap="Original Model"}
# Original model plots
par(mfrow = c(1,3))
plot(fit.lin, se = TRUE, all.terms = TRUE)
``` 

```{r model2Plot, fig.width=8, fig.cap="Polynomial Model"}
# Polynomial model plots
par(mfrow = c(1,3))
plot(fit.poly, se = TRUE, all.terms = TRUE)
```

```{r model3Plot, fig.width=8, fig.cap="Natural Splines Model"}
# Natural splines model plots
par(mfrow = c(1,3))
plot(fit.gamns, se = TRUE, all.terms = TRUE)
```


```{r model4Plot, fig.width=8, fig.cap="Combined Model"}
# Combined model plots
par(mfrow = c(1,3))
plot(fit.gammix, se = TRUE, all.terms = TRUE)
```

```{r model5Plot, fig.width=8, fig.cap="Natural Splines Model with different predictors"}
# Different predictors model plots
par(mfrow = c(1,3))
plot(fit.gamnsDiff, se = TRUE, all.terms = TRUE)
```

```{r predSelect, fig.cap="Subset Selection of Predictors", fig.height=4, fig.width=6}
# Selection of predictors
plot(summary(regfit.full)$rss, 
     xlab="Number of Predictors", 
     ylab="Residual Sum of Squares", 
     type="b",
     col='red')
``` 

\pagebreak 

## R Code

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```


